Social Participation Ontology: community documentation, enhancements and use examples

### 2015-09-22 plosOne/plos.tex
terminado de revisar todo artigo

\section{VCPS original documentation}
From April to December, 2012, \vcps\ was first conceived. In the online process, as registered by \corais\ platform, 66 users interacted, 6 of them were the most active~\cite{metodologia}. Various materials were produced both as activity traces and as reference media. This section is dedicated to these materials.

@@

\section{Concluding remarks}\label{conc}
\opsi, based in \vcps,
yields initial steps in achieving an effective social participation ontology: community has registered activities and delivered reference documents, including this article.
Sections~\ref{exp} and~\ref{ospUtil} have expansion directions for the ontology, regarding uses, further definition of classes, upper ontologies and related ontological structures.

@@

\subsection{Further work}
Further work involves observing community manifestations about \ops\ and this article, accomplishing use by means of formal instances and civil society, and studying upper ontologies usage with \ops. The use of \ops\ (or a variant) in different instances is being tackled for the creation of social participation indicators and easing participation processes.

Academic texts dedicated to each new participation ontology created (\ocd, \obs, \ontologiaa, \opa) should be written and submitted to peer review for enhancements and quality assurance~\cite{pnud5}.

Data related to these ontologies were found in relational databases and preliminary scripts were written to make them available as \rdf. A sound linkage of this data and incorporation to the Linked Open Data (\lod~\cite{lod}) cloud is planed for a near future. This should make Brazilian participative structures and data part of the Giant Global Graph~\cite{ggg}.

\subsubsection*{Acknowledgments}
This article is deeply influenced by Protege and BFO documentation. Authors thank community and researchers related to these projects~\cite{protege,bfo}.
Authors thank \corais\ Platform maintainers for their efforts in delivering a collaborative platform which gave birth to \vcps\ documentation.
Renato Fabbri is grateful to CNPq (process 140860/2013-4, project 870336/1997-5), PNUD/ONU (contract 2013/00056, project BRA/12/018), SNAS/SG-PR, and the Postgraduate Committee of the IFSC/USP. 
% Authors thanks Prof. Dr. Dilvan de Abreu Moreira (ICMC/USP) for all help and incentives in the genesis of this article. Authors thank Flor Karina Mamani Amanqui for initial Spanish labels.

@@

Present in preliminary \ops\ implementation, all restrictions were removed from current \ops\ as explained in Section~\ref{impl}.
\begin{table}[!h]
  \centering
  \caption{Restrictions of the \ops\ (Ontology of Social Participation). All \ops\ restrictions are existential (\texttt{owl:someValuesFrom)}.}
  \begin{tabular}{|l|l|l|}\hline
{\bf subject} & {\bf predicate} & {\bf object} \\\hline\hline
Initiator    & starts        & Cause\\\hline
Supporter    & supports      & Cause\\\hline
Executer     & executes     & Action\\\hline
Solution     & solves      & Problem\\\hline
SocialActor  & hasRole            & Role\\\hline
Action       & produces    & Results\\\hline
Result       & contributesTo   & Solution\\\hline
Cause        & action          & Action\\\hline
Action       & belongsTo     & Scope\\\hline
Cause        & theme           & Theme\\\hline
Cause        & proposes   & Solution\\\hline
Problem      & generates     & Cause\\\hline
  \end{tabular}
  \label{ospRestr}
\end{table}

>>>>

\section{VCPS original documentation}\label{sec:orig}
From April to December, 2012, \vcps\ was first conceived. In the online process, as registered by \corais\ platform, 66 users interacted, 6 of them were the most active~\cite{metodologia}. Various materials were produced both as activity traces and as reference media. This section is dedicated to these materials.

@@

\section{Concluding remarks and further work}\label{conc}
\opsi, based in \vcps,
yields initial steps in achieving an effective social participation ontology.
Community has registered activities and delivered reference documents, including this present article which contains 
an organization of the community foundational conceptualizations,
\ops\ diligent development, uses,
upper ontologies and expansion of the ontological structures.

@@

%\subsection{Further work}
Further work involves observing community manifestations about \ops\ and this article, accomplishing use by means of formal instances and civil society, and studying upper ontologies usage with \ops. The use of \ops\ (or a variant) in different instances is being tackled for the creation of social participation indicators and easing participation processes.
Academic texts dedicated to the new participation ontologies created (\ocd, \obs, \ontologiaa, \opa) should be written and submitted to peer review for enhancements and quality assurance~\cite{pnud5}.

Data related to these ontologies were found in relational databases and preliminary scripts were written to make them available as \rdf~\cite{datahub}. A sound linkage of this data and the consequent incorporation to the Linked Open Data (\lod~\cite{lod}) cloud is planed for a near future. This should make Brazilian participative structures and data more relevant to the Giant Global Graph~\cite{ggg}.

\subsubsection*{Acknowledgments}
This article is deeply influenced by \protege\ and \bfo\ documentation. Authors thank community and researchers related to these projects~\cite{protege,bfo}.
Authors thank \corais\ platform maintainers for their efforts in delivering a collaborative platform which gave birth to \vcps\ documentation.
Renato Fabbri is grateful to \textsc{cnp}q (process 140860/2013-4, project 870336/1997-5), \pnud\ (contract 2013/00056, project BRA/12/018), \textsc{snas/sg-pr}, and the postgraduate committee of the \textsc{ifsc/usp}.
% Authors thanks Prof. Dr. Dilvan de Abreu Moreira (ICMC/USP) for all help and incentives in the genesis of this article. Authors thank Flor Karina Mamani Amanqui for initial Spanish labels.

@@

Present in preliminary \ops\ implementation, all restrictions were removed from current \ops\ as explained in Section~\ref{impl}.
\begin{table}[!h]
  \centering
  \caption{Restrictions of the preliminary \ops: all restrictions are existential (\texttt{owl:someValuesFrom)}.}
  \begin{tabular}{|l|l|l|}\hline
{\bf subject} & {\bf predicate} & {\bf object} \\\hline\hline
Initiator    & starts        & Cause\\\hline
Supporter    & supports      & Cause\\\hline
Executer     & executes     & Action\\\hline
Solution     & solves      & Problem\\\hline
SocialActor  & hasRole            & Role\\\hline
Action       & produces    & Results\\\hline
Result       & contributesTo   & Solution\\\hline
Cause        & action          & Action\\\hline
Action       & belongsTo     & Scope\\\hline
Cause        & theme           & Theme\\\hline
Cause        & proposes   & Solution\\\hline
Problem      & generates     & Cause\\\hline
  \end{tabular}
  \label{ospRestr}
\end{table}
